---
title: "Structural Phylogenomics Pipeline Tutorial"
subtitle: "From Bacterial Genomes to Phylogenetic Trees"
author: "Your Name"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    toc-title: "Pipeline Steps"
    code-fold: false
    theme: cosmo
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This tutorial walks through a structural phylogenomics pipeline step by step,
using the *E. coli* K12 genome (GenBank: U00096.3) as our example dataset and
**MotA** (flagellar motor protein, UniProt: P0AFN2) as our query sequence and
structure.

Each section covers one stage of the pipeline. Within each section you will:

1. **Check** that the required tool is installed and working
2. **Run** a simple example using the tutorial data

Work through the sections in order. All input data are pre-staged in the
numbered folders alongside this file.

---

# R packages

Load all required R packages here. If any are missing, install them with
`install.packages("pak"); pak::pkg_install(c(...))`.

```{r}
#| label: load-packages

library(Biostrings)  # FASTA handling
library(ape)         # tree reading and manipulation
library(ggtree)      # tree visualization
library(ggplot2)     # general plotting
library(TreeDist)    # tree distance statistics
library(tidyverse)   # data wrangling
```

---

# Step 01 — Input Data

**Folder:** `01_input/`

This step has no external tool dependency. We inspect the input data to make
sure everything is in order before we start.

## Input files

| File | Description |
|---|---|
| `01_input/ecoli_k12.gbk` | *E. coli* K12 MG1655 genome (GenBank format, U00096.3) |
| `01_input/motA.faa` | MotA query sequence (FASTA amino acid, P0AFN2) |
| `01_input/motA.pdb` | MotA AlphaFold structure (PDB format, AF-P0AFN2-F1) |

## Check: inspect the genome file

```{bash}
#| label: check-genome

echo "=== Genome file (first 5 lines) ==="
head -5 01_input/ecoli_k12.gbk

echo ""
echo "=== Total lines in genome file ==="
wc -l 01_input/ecoli_k12.gbk
```

## Check: inspect the MotA query sequence

```{r}
#| label: check-motA-fasta

motA <- readAAStringSet("01_input/motA.faa")
motA
cat("\nSequence length:", width(motA), "amino acids\n")
```

## Check: inspect the MotA AlphaFold structure

```{bash}
#| label: check-motA-pdb

echo "=== PDB file header ==="
grep "^HEADER\|^TITLE\|^SEQRES" 01_input/motA.pdb | head -10

echo ""
echo "=== Number of ATOM records ==="
grep -c "^ATOM" 01_input/motA.pdb
```

---

# Step 02 — Extract Proteins from Genome

**Folder:** `02_genome_proteins/`  
**Tool:** standard Unix tools (`grep`, `awk`) + R (Biostrings)

Before searching for homologs, we need to extract all protein sequences from
the GenBank genome file into a FASTA file.

## Check: confirm R and Biostrings are working

```{r}
#| label: check-biostrings

packageVersion("Biostrings")
```

## Run: extract proteins from GenBank file

We use the `readGenBank()` approach via the `genbankr` package, or
alternatively parse the GenBank file for CDS translation fields using
Biostrings. Here we use a shell approach with `any2fasta` or direct GenBank
parsing — confirm which tool you have available:

```{bash}
#| label: check-any2fasta

any2fasta --version 2>&1 || echo "any2fasta not found — see note below"
```

> **Note:** If `any2fasta` is not installed, install it with:
> `conda install -c bioconda any2fasta` or `brew install any2fasta`

```{bash}
#| label: extract-proteins

any2fasta 01_input/ecoli_k12.gbk > 02_genome_proteins/ecoli_k12_proteins.faa

echo "=== Extraction complete. Sequence count: ==="
grep -c ">" 02_genome_proteins/ecoli_k12_proteins.faa

echo ""
echo "=== First 3 sequences: ==="
grep ">" 02_genome_proteins/ecoli_k12_proteins.faa | head -3
```

## Check: verify MotA is present in the extracted proteins

```{r}
#| label: verify-motA-present

proteins <- readAAStringSet("02_genome_proteins/ecoli_k12_proteins.faa")
cat("Total proteins extracted:", length(proteins), "\n")

# Search for MotA by name in headers
mota_hit <- proteins[grepl("motA|MotA|P0AFN2", names(proteins))]
cat("MotA entries found:", length(mota_hit), "\n")
print(mota_hit)
```

---

# Step 03 — Homolog Search with HMMer

**Folder:** `03_hmmer/`  
**Tool:** [HMMer](http://hmmer.org/) (`hmmbuild`, `hmmsearch`)

HMMer searches for sequence homologs using profile hidden Markov models
(HMMs). We will build a profile from our single MotA query, then search it
against the *E. coli* K12 proteome as a test.

## Check: confirm HMMer is installed

```{bash}
#| label: check-hmmer

hmmbuild -h 2>&1 | head -3
echo "---"
hmmsearch -h 2>&1 | head -3
```

## Run: build an HMM profile from MotA

A single-sequence "profile" is the simplest case. In a real run you would
build from a curated multiple sequence alignment.

```{bash}
#| label: hmmbuild

hmmbuild \
  03_hmmer/motA.hmm \
  01_input/motA.faa

echo "=== HMM profile created ==="
grep "^NAME\|^LENG\|^ALPH" 03_hmmer/motA.hmm
```

## Run: search the profile against the K12 proteome

```{bash}
#| label: hmmsearch

hmmsearch \
  --tblout 03_hmmer/hits.tbl \
  --cpu 4 \
  -E 1e-5 \
  03_hmmer/motA.hmm \
  02_genome_proteins/ecoli_k12_proteins.faa

echo "=== Hits (excluding comment lines) ==="
grep -v "^#" 03_hmmer/hits.tbl | head -10
```

## Inspect HMMer results in R

```{r}
#| label: hmmer-results

hmmer_hits <- read.table(
  "03_hmmer/hits.tbl",
  comment.char = "#",
  header        = FALSE,
  fill          = TRUE
) |>
  select(target = V1, evalue = V5, score = V6) |>
  mutate(evalue = as.numeric(evalue), score = as.numeric(score)) |>
  arrange(evalue)

cat("Total hits:", nrow(hmmer_hits), "\n\n")
print(hmmer_hits)
```

---

# Step 04 — Homolog Search with Foldseek

**Folder:** `04_foldseek/`  
**Tool:** [Foldseek](https://github.com/steinegger-lab/foldseek)

Foldseek searches for structural homologs by comparing protein structures
directly. We search the MotA AlphaFold structure against a small local
database built from the *E. coli* K12 AlphaFold structures.

## Check: confirm Foldseek is installed

```{bash}
#| label: check-foldseek

foldseek version
```

## Run: build a Foldseek database from the query structure

```{bash}
#| label: foldseek-createdb

# Create a small database from the query structure itself (as a self-search test)
mkdir -p 04_foldseek/query_db
foldseek createdb 01_input/motA.pdb 04_foldseek/query_db/motA_db

echo "=== Query database created ==="
ls 04_foldseek/query_db/
```

## Run: search MotA structure against the query database (self-search test)

In the full pipeline you would search against a database of all *E. coli*
AlphaFold structures. Here we run a self-search to confirm Foldseek is
working — MotA should hit itself with a perfect score.

```{bash}
#| label: foldseek-search

foldseek easy-search \
  01_input/motA.pdb \
  04_foldseek/query_db/motA_db \
  04_foldseek/hits.tsv \
  04_foldseek/tmp \
  --format-output "query,target,evalue,bits,lddt,prob" \
  -e 1

echo "=== Foldseek hits ==="
cat 04_foldseek/hits.tsv
```

---

# Step 05 — Extract Hits to FASTA

**Folder:** `05_hit_fasta/`  
**Tool:** R (Biostrings)

We combine the HMMer and Foldseek hit lists and extract the corresponding
sequences from the proteome FASTA file.

## Run: extract hit sequences

```{r}
#| label: extract-hits

library(Biostrings)

# Load all proteins
proteins <- readAAStringSet("02_genome_proteins/ecoli_k12_proteins.faa")

# Collect hit IDs from HMMer (already loaded above)
hit_ids <- hmmer_hits$target

# Match against protein headers
# (headers may contain the ID as part of a longer string)
matched <- proteins[sapply(names(proteins), function(nm) {
  any(sapply(hit_ids, function(id) grepl(id, nm, fixed = TRUE)))
})]

# Always include the query itself
query <- readAAStringSet("01_input/motA.faa")
hits_with_query <- c(query, matched)

cat("Query sequence:      1\n")
cat("HMMer hits matched:", length(matched), "\n")
cat("Total in output:   ", length(hits_with_query), "\n\n")

writeXStringSet(hits_with_query, "05_hit_fasta/hits.faa")
cat("Written to 05_hit_fasta/hits.faa\n")
```

## Check: inspect the output FASTA

```{bash}
#| label: check-hit-fasta

echo "=== Sequences in hits.faa ==="
grep ">" 05_hit_fasta/hits.faa

echo ""
echo "=== First sequence ==="
head -4 05_hit_fasta/hits.faa
```

---

# Step 06 — Convert Structures to 3Di with Foldseek

**Folder:** `06_3di/`  
**Tool:** Foldseek (`createdb`, `convert2fasta`)

Foldseek encodes protein structures as sequences of structural "letters" using
the 3Di alphabet — one character per residue. This lets us align structures
as if they were sequences.

## Check: foldseek is available (already checked in Step 04)

```{bash}
#| label: check-foldseek-2

foldseek version
```

## Run: convert MotA structure to 3Di sequence

```{bash}
#| label: struct-to-3di

mkdir -p 06_3di/db

# Create a Foldseek database from the AlphaFold structure
foldseek createdb 01_input/motA.pdb 06_3di/db/motA_db

# Convert to FASTA with 3Di characters
foldseek convert2fasta 06_3di/db/motA_db 06_3di/motA_3di.fasta

echo "=== 3Di sequence for MotA ==="
cat 06_3di/motA_3di.fasta
```

## Compare AA vs 3Di sequences in R

```{r}
#| label: compare-aa-3di

aa_seq <- readAAStringSet("01_input/motA.faa")
di_seq <- readAAStringSet("06_3di/motA_3di.fasta")

cat("AA  length:", width(aa_seq), "\n")
cat("3Di length:", width(di_seq), "\n\n")

cat("AA  (first 60):", substr(as.character(aa_seq[[1]]),  1, 60), "\n")
cat("3Di (first 60):", substr(as.character(di_seq[[1]]),  1, 60), "\n")
```

---

# Step 07 — Multiple Sequence Alignment

**Folder:** `07_alignments/`  
**Tools:** [MAFFT](https://mafft.cbrc.jp/), [FAMSA](https://github.com/refresh-bio/FAMSA), [Foldmason](https://github.com/steinegger-lab/foldmason)

We align the hit sequences using three complementary methods.

## 7a: MAFFT (amino acid alignment)

### Check

```{bash}
#| label: check-mafft

mafft --version 2>&1
```

### Run

```{bash}
#| label: run-mafft

mafft --auto --thread 4 05_hit_fasta/hits.faa \
  > 07_alignments/hits_mafft_aa.fasta

echo "=== MAFFT alignment ==="
grep -c ">" 07_alignments/hits_mafft_aa.fasta
echo "sequences aligned"
grep -v ">" 07_alignments/hits_mafft_aa.fasta | head -2
```

## 7b: FAMSA with 3Di sequences

### Check

```{bash}
#| label: check-famsa

famsa --version 2>&1 | head -2
```

### Run

> **Note:** In the full pipeline you would align 3Di sequences for all hits.
> Here we use only the MotA 3Di sequence as a single-sequence demonstration,
> since our *E. coli* example has few hits. With more taxa, pass a multi-FASTA
> 3Di file instead.

```{bash}
#| label: run-famsa

famsa \
  -t 4 \
  06_3di/motA_3di.fasta \
  07_alignments/hits_famsa_3di.fasta

echo "=== FAMSA 3Di alignment ==="
cat 07_alignments/hits_famsa_3di.fasta
```

## 7c: Foldmason (structure-guided alignment)

### Check

```{bash}
#| label: check-foldmason

foldmason version 2>&1 | head -2
```

### Run

```{bash}
#| label: run-foldmason

foldmason easy-msa \
  01_input/ \
  07_alignments/hits_foldmason \
  07_alignments/foldmason_tmp \
  --report-mode 1

echo "=== Foldmason output files ==="
ls 07_alignments/ | grep foldmason
```

## Alignment summary in R

```{r}
#| label: alignment-summary

library(Biostrings)

aln_mafft <- readAAStringSet("07_alignments/hits_mafft_aa.fasta")

cat("MAFFT AA alignment\n")
cat("  Sequences:       ", length(aln_mafft), "\n")
cat("  Alignment width: ", unique(width(aln_mafft)), "columns\n")

gap_freq <- colMeans(as.matrix(aln_mafft) == "-")
cat("  Mean gap freq:   ", round(mean(gap_freq), 3), "\n")
```

## Gap profile plot

```{r}
#| label: gap-plot
#| fig-width: 10
#| fig-height: 3

tibble(position = seq_along(gap_freq), gap_freq = gap_freq) |>
  ggplot(aes(x = position, y = gap_freq)) +
  geom_area(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Gap frequency across MAFFT AA alignment",
    x     = "Alignment position",
    y     = "Proportion of gaps"
  ) +
  theme_minimal()
```

---

# Step 08 — Phylogenetic Inference with IQ-TREE 2

**Folder:** `08_trees/`  
**Tool:** [IQ-TREE 2](http://www.iqtree.org/)

IQ-TREE 2 infers maximum-likelihood phylogenetic trees. We run three
analyses: AA-only, 3Di-only, and a partitioned AA+3Di dataset.

## Check: confirm IQ-TREE 2 is installed

```{bash}
#| label: check-iqtree

iqtree2 --version 2>&1 | head -3
```

## Run: AA-only tree

```{bash}
#| label: iqtree-aa

iqtree2 \
  -s 07_alignments/hits_mafft_aa.fasta \
  --prefix 08_trees/tree_aa \
  -m TEST \
  -B 1000 \
  -T AUTO \
  --redo 2>&1 | tail -5

echo "=== Output files ==="
ls 08_trees/tree_aa.*
```

## Run: 3Di-only tree

```{bash}
#| label: iqtree-3di

iqtree2 \
  -s 07_alignments/hits_famsa_3di.fasta \
  --prefix 08_trees/tree_3di \
  -m TEST \
  -B 1000 \
  -T AUTO \
  --redo 2>&1 | tail -5

echo "=== Output files ==="
ls 08_trees/tree_3di.*
```

## Run: partitioned AA + 3Di tree

```{bash}
#| label: iqtree-partition

cat > 08_trees/partition.nex << 'EOF'
#nexus
begin sets;
  charset AA  = 07_alignments/hits_mafft_aa.fasta: *;
  charset 3Di = 07_alignments/hits_famsa_3di.fasta: *;
end;
EOF

iqtree2 \
  -p 08_trees/partition.nex \
  --prefix 08_trees/tree_partitioned \
  -B 1000 \
  -T AUTO \
  --redo 2>&1 | tail -5

echo "=== Output files ==="
ls 08_trees/tree_partitioned.*
```

---

# Step 09 — Tree Visualization in R

**Folder:** `09_visualization/`  
**Tools:** R (`ggtree`, `ape`), FigTree

## Check: confirm ggtree is available

```{r}
#| label: check-ggtree

packageVersion("ggtree")
packageVersion("ape")
```

## Run: plot the AA tree

```{r}
#| label: plot-aa-tree
#| fig-width: 10
#| fig-height: 6

library(ape)
library(ggtree)

tree_aa <- read.tree("08_trees/tree_aa.treefile")

ggtree(tree_aa) +
  geom_tiplab(size = 3.5, offset = 0.01) +
  geom_nodelab(
    aes(label = ifelse(as.numeric(label) >= 70, label, "")),
    size  = 2.5,
    color = "firebrick",
    hjust = 1.2
  ) +
  labs(title = "AA tree — MAFFT + IQ-TREE 2") +
  theme_tree2() +
  xlim(0, max(node.depth.edgelength(tree_aa)) * 1.3)
```

## Run: plot the 3Di tree

```{r}
#| label: plot-3di-tree
#| fig-width: 10
#| fig-height: 6

tree_3di <- read.tree("08_trees/tree_3di.treefile")

ggtree(tree_3di) +
  geom_tiplab(size = 3.5, offset = 0.01) +
  geom_nodelab(
    aes(label = ifelse(as.numeric(label) >= 70, label, "")),
    size  = 2.5,
    color = "steelblue",
    hjust = 1.2
  ) +
  labs(title = "3Di tree — FAMSA + IQ-TREE 2") +
  theme_tree2() +
  xlim(0, max(node.depth.edgelength(tree_3di)) * 1.3)
```

> **Interactive exploration:** Open the `.treefile` outputs in **FigTree** for
> interactive re-rooting, colouring, and annotation.  
> Launch FigTree and use *File → Open* to load `08_trees/tree_aa.treefile`.

---

# Step 10 — Tree Distance Statistics

**Folder:** `10_tree_distances/`  
**Tool:** R (`TreeDist`, `ape`)

When two proteins are hypothesized to share vertical (co-speciating)
evolutionary history, their gene trees should be congruent. We quantify tree
similarity using several distance metrics.

## Check: confirm TreeDist is available

```{r}
#| label: check-treedist

packageVersion("TreeDist")
```

## Run: calculate tree distances

```{r}
#| label: tree-distances

library(TreeDist)
library(ape)

tree_aa  <- read.tree("08_trees/tree_aa.treefile")
tree_3di <- read.tree("08_trees/tree_3di.treefile")

# Subset to shared tip labels
shared <- intersect(tree_aa$tip.label, tree_3di$tip.label)
cat("Shared tips:", length(shared), "\n\n")

t1 <- keep.tip(tree_aa,  shared)
t2 <- keep.tip(tree_3di, shared)

distances <- tibble(
  metric = c(
    "Robinson-Foulds distance",
    "Clustering Info distance",
    "Matching Split distance",
    "Nye similarity (0-1)"
  ),
  value = c(
    RobinsonFoulds(t1, t2),
    ClusteringInfoDistance(t1, t2),
    MatchingSplitDistance(t1, t2),
    NyeSimilarity(t1, t2)
  )
)

print(distances)
```

## Interpreting the metrics

| Metric | Range | Identical trees | Completely different |
|---|---|---|---|
| Robinson-Foulds | 0 – 2(n-3) | 0 | maximum |
| Clustering Info | 0 – ... | 0 | large |
| Matching Split | 0 – ... | 0 | large |
| Nye similarity | 0 – 1 | 1 | 0 |

A low Robinson-Foulds distance combined with a high Nye similarity suggests
the two trees (and therefore the two proteins) share the same evolutionary
history — consistent with vertical inheritance.

---

# Session Info

```{r}
#| label: session-info
sessionInfo()
```
