---
title: "Structural Phylogenomics Pipeline"
subtitle: "From Bacterial Genomes to Phylogenetic Trees"
author: "Your Name"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    toc-title: "Pipeline Steps"
    code-fold: false
    theme: cosmo
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This tutorial walks through a structural phylogenomics pipeline applied to a
sample of bacterial genomes. We will search for homologs using sequence and
structure-based methods, retrieve AlphaFold structural predictions, construct
multiple sequence alignments incorporating both amino acid (AA) and structural
(3Di) information, infer phylogenetic trees, and assess tree topology and
distance.

**Prerequisites:** Before starting, ensure the following tools are installed
and available on your `PATH`:

- [HMMer](http://hmmer.org/) (`hmmbuild`, `hmmsearch`)
- [Foldseek](https://github.com/steinegger-lab/foldseek)
- [MAFFT](https://mafft.cbrc.jp/alignment/software/)
- [FAMSA](https://github.com/refresh-bio/FAMSA) (with 3Di support)
- [Foldmason](https://github.com/steinegger-lab/foldmason)
- [IQ-TREE 2](http://www.iqtree.org/)
- [FigTree](http://tree.bio.ed.ac.uk/software/figtree/) (for interactive tree viewing)

The following R packages are also required:

```{r}
#| label: load-packages

# Install any missing packages with: install.packages("pak"); pak::pkg_install(...)
library(ape)        # tree reading and manipulation
library(ggtree)     # tree visualization
library(ggplot2)    # general plotting
library(Biostrings) # FASTA handling
library(TreeDist)   # tree distance statistics
library(tidyverse)  # data wrangling
```

---

# Step 1: Define Your Input Data

Set your working paths here. All subsequent steps reference these variables,
so you only need to update this block if your file locations change.

```{bash}
#| label: setup-dirs

# Create output directories if they don't exist
mkdir -p results/hmmer
mkdir -p results/foldseek
mkdir -p results/fasta
mkdir -p results/alphafold
mkdir -p results/3di
mkdir -p results/alignments
mkdir -p results/trees

echo "Directory structure ready."
ls results/
```

---

# Step 2: Homolog Search

## 2a: Profile-based search with HMMer

Use `hmmsearch` to search your bacterial genome protein sequences against an
HMM profile. Replace `my_profile.hmm` and `genome_proteins.faa` with your
actual file names.

```{bash}
#| label: hmmsearch

hmmsearch \
  --tblout results/hmmer/hits.tbl \
  --cpu 4 \
  -E 1e-5 \
  data/profiles/my_profile.hmm \
  data/raw/genome_proteins.faa

echo "HMMer search complete. Top hits:"
head -20 results/hmmer/hits.tbl | grep -v "^#"
```

## 2b: Structure-based search with Foldseek

Use `foldseek easy-search` to search with a query structure against a database
of bacterial proteome structures.

```{bash}
#| label: foldseek-search

foldseek easy-search \
  data/query_structures/ \
  data/foldseek_db/bacterial_db \
  results/foldseek/hits.tsv \
  results/foldseek/tmp \
  --exhaustive-search \
  --format-output "query,target,evalue,bits,lddt,prob" \
  -e 1e-5

echo "Foldseek search complete. Top hits:"
head -20 results/foldseek/hits.tsv
```

---

# Step 3: Extract Hits to FASTA

Parse the HMMer and/or Foldseek output to extract hit sequences into a FASTA
file. Here we use R for flexible parsing.

```{r}
#| label: extract-hits

library(Biostrings)

# --- Parse HMMer hits table ---
hmmer_hits <- read.table(
  "results/hmmer/hits.tbl",
  comment.char = "#",
  header = FALSE
) |>
  select(target = V1, evalue = V5) |>
  filter(as.numeric(evalue) < 1e-5)

cat("HMMer hits passing threshold:", nrow(hmmer_hits), "\n")
print(head(hmmer_hits))

# --- Load genome proteins and extract hits ---
all_seqs <- readAAStringSet("data/raw/genome_proteins.faa")

# Match sequence names to hit IDs
hit_names <- hmmer_hits$target
hit_seqs  <- all_seqs[names(all_seqs) %in% hit_names]

# Write hits to FASTA
writeXStringSet(hit_seqs, "results/fasta/hits.faa")
cat("\nWrote", length(hit_seqs), "sequences to results/fasta/hits.faa\n")
```

---

# Step 4: Retrieve AlphaFold Structures

For each hit, download its AlphaFold predicted structure from the EBI
AlphaFold database. The accession is assumed to be a UniProt ID.

```{bash}
#| label: alphafold-download

# Extract UniProt IDs from the FASTA headers (adjust awk pattern to your headers)
grep ">" results/fasta/hits.faa | awk '{print $1}' | sed 's/>//' \
  > results/alphafold/uniprot_ids.txt

echo "Downloading AlphaFold structures for:"
cat results/alphafold/uniprot_ids.txt

# Download each structure
while read accession; do
  url="https://alphafold.ebi.ac.uk/files/AF-${accession}-F1-model_v4.pdb"
  wget -q -O "results/alphafold/${accession}.pdb" "$url" \
    && echo "  Downloaded: $accession" \
    || echo "  Not found:  $accession"
done < results/alphafold/uniprot_ids.txt
```

---

# Step 5: Convert Structures to 3Di with Foldseek

Foldseek encodes protein structures as sequences of structural "letters" (3Di
alphabet). We extract these using `foldseek structureto3didescriptor`.

```{bash}
#| label: struct-to-3di

foldseek createdb results/alphafold/ results/3di/af_db
foldseek convert2fasta results/3di/af_db results/3di/hits_3di.fasta --db-load-mode 3

echo "3Di FASTA created:"
grep -c ">" results/3di/hits_3di.fasta
```

---

# Step 6: Multiple Sequence Alignment

We generate alignments using three methods: standard AA alignment (MAFFT),
3Di-aware alignment (FAMSA with 3Di), and structure-guided alignment
(Foldmason). These can be compared in Step 7.

## 6a: MAFFT (amino acid)

```{bash}
#| label: mafft

mafft --auto --thread 4 results/fasta/hits.faa \
  > results/alignments/hits_mafft_aa.fasta

echo "MAFFT alignment complete."
grep -c ">" results/alignments/hits_mafft_aa.fasta
```

## 6b: FAMSA with 3Di

```{bash}
#| label: famsa-3di

famsa \
  -t 4 \
  results/3di/hits_3di.fasta \
  results/alignments/hits_famsa_3di.fasta

echo "FAMSA 3Di alignment complete."
grep -c ">" results/alignments/hits_famsa_3di.fasta
```

## 6c: Foldmason (structure-guided)

```{bash}
#| label: foldmason

foldmason easy-msa \
  results/alphafold/ \
  results/alignments/hits_foldmason \
  results/alignments/foldmason_tmp \
  --report-mode 1

echo "Foldmason alignment complete."
```

---

# Step 7: Assess Alignments

## 7a: Summary statistics in R

```{r}
#| label: alignment-stats

library(Biostrings)
library(tidyverse)

# Read alignments
alns <- list(
  mafft_aa  = readAAStringSet("results/alignments/hits_mafft_aa.fasta"),
  famsa_3di = readAAStringSet("results/alignments/hits_famsa_3di.fasta")
)

# Summarize each alignment
aln_summary <- map_dfr(names(alns), function(nm) {
  aln <- alns[[nm]]
  tibble(
    method    = nm,
    n_seqs    = length(aln),
    aln_width = unique(width(aln)),
    pct_gap   = round(
      mean(letterFrequency(aln, "-") / width(aln)) * 100, 2
    )
  )
})

print(aln_summary)
```

## 7b: Alignment visualization

```{r}
#| label: alignment-plot
#| fig-width: 10
#| fig-height: 4

# Simple gap-density plot across alignment columns for MAFFT alignment
aln_mat <- as.matrix(alns$mafft_aa)
gap_freq <- colMeans(aln_mat == "-")

tibble(position = seq_along(gap_freq), gap_freq = gap_freq) |>
  ggplot(aes(x = position, y = gap_freq)) +
  geom_area(fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Gap frequency across MAFFT AA alignment",
    x     = "Alignment position",
    y     = "Proportion of gaps"
  ) +
  theme_minimal()
```

---

# Step 8: Phylogenetic Inference with IQ-TREE 2

We run three tree searches: AA-only, 3Di-only, and a partitioned AA+3Di
analysis.

## 8a: AA-only tree

```{bash}
#| label: iqtree-aa

iqtree2 \
  -s results/alignments/hits_mafft_aa.fasta \
  --prefix results/trees/tree_aa \
  -m TEST \
  -B 1000 \
  -T AUTO \
  --redo

echo "IQ-TREE AA tree complete."
```

## 8b: 3Di-only tree

```{bash}
#| label: iqtree-3di

iqtree2 \
  -s results/alignments/hits_famsa_3di.fasta \
  --prefix results/trees/tree_3di \
  -m TEST \
  -B 1000 \
  -T AUTO \
  --redo

echo "IQ-TREE 3Di tree complete."
```

## 8c: Partitioned AA + 3Di tree

```{bash}
#| label: iqtree-partitioned

# Create partition file
cat > results/trees/partition.nex << 'EOF'
#nexus
begin sets;
  charset AA  = results/alignments/hits_mafft_aa.fasta: *;
  charset 3Di = results/alignments/hits_famsa_3di.fasta: *;
end;
EOF

iqtree2 \
  -p results/trees/partition.nex \
  --prefix results/trees/tree_partitioned \
  -B 1000 \
  -T AUTO \
  --redo

echo "IQ-TREE partitioned tree complete."
```

---

# Step 9: Tree Visualization in R

```{r}
#| label: tree-viz
#| fig-width: 10
#| fig-height: 8

library(ape)
library(ggtree)

# Read trees
tree_aa   <- read.tree("results/trees/tree_aa.treefile")
tree_3di  <- read.tree("results/trees/tree_3di.treefile")

# Plot AA tree with bootstrap support
p_aa <- ggtree(tree_aa) +
  geom_tiplab(size = 3) +
  geom_nodelab(aes(label = label), size = 2.5, color = "firebrick") +
  labs(title = "AA tree (MAFFT + IQ-TREE)") +
  theme_tree2()

print(p_aa)

# Plot 3Di tree
p_3di <- ggtree(tree_3di) +
  geom_tiplab(size = 3) +
  geom_nodelab(aes(label = label), size = 2.5, color = "steelblue") +
  labs(title = "3Di tree (FAMSA + IQ-TREE)") +
  theme_tree2()

print(p_3di)
```

> **Note:** For interactive exploration, open the `.treefile` outputs in
> FigTree, which allows you to re-root, colour, and annotate trees manually.

---

# Step 10: Tree Distance Statistics

When two proteins are hypothesized to share vertical evolutionary history, we
can quantify how similar their trees are using tree distance metrics.

```{r}
#| label: tree-distances

library(TreeDist)

# Make sure tip labels match between trees before comparing
# (subset to shared tips if needed)
shared_tips <- intersect(tree_aa$tip.label, tree_3di$tip.label)
tree_aa_sub  <- keep.tip(tree_aa,  shared_tips)
tree_3di_sub <- keep.tip(tree_3di, shared_tips)

# Calculate several distance metrics
distances <- tibble(
  metric = c(
    "Robinson-Foulds",
    "Clustering Info Distance",
    "Matching Split Distance",
    "Nye Similarity"
  ),
  value = c(
    RobinsonFoulds(tree_aa_sub, tree_3di_sub),
    ClusteringInfoDistance(tree_aa_sub, tree_3di_sub),
    MatchingSplitDistance(tree_aa_sub, tree_3di_sub),
    NyeSimilarity(tree_aa_sub, tree_3di_sub)
  )
)

print(distances)
```

Interpretation guide:

- **Robinson-Foulds distance**: counts differing bipartitions; 0 = identical topologies.
- **Clustering Info Distance**: information-theoretic; lower = more similar.
- **Matching Split Distance**: accounts for partial agreement between splits.
- **Nye Similarity**: 0–1 similarity score; higher = more congruent trees.

---

# Session Info

```{r}
#| label: session-info

sessionInfo()
```
